Modelling state-switching functional data with hidden Markov models

### 2020-12-17 Report/Sections/models.tex
Updates from Overleaf

Consider a sequence of $T$ curves, where any particular curve $t$ is characterized by a curve-level (or coarse-scale) observation $Y_t$ as well as a sequence of $T^*_t$ within-curve (or fine-scale) observations $Y^*_{t}$. Namely, $Y^*_{t} \equiv \big\{Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\big\}$ is made up of function values from curve $t$. We call the sequence of coarse-scale observations $Y \equiv \big\{Y_1, \ldots, Y_T\big\}$ and the collection of all fine-scale observations $Y^* \equiv \big\{Y^*_1,\ldots,Y^*_T \big\}$. We first detail the structure of a traditional HMM followed by three variations which generalize its base structure. We then show how each of these generalized HMMs can be synthesized to form a wide variety of more complicated models.

@@

Using observation emissions, denoted here as $y \equiv \{y_1,\ldots,y_T\}$, we can find the maximum likelihood estimates of the parameters $\Gamma$ and $\theta \equiv \{\theta^{(1)},\ldots,\theta^{(N)}\}$. The likelihood $\calL_{\text{HMM}}$ can be evaluated using well-known \textit{forward algorithm} \citep{Zucchini:2016}:
%
$$\calL_{\text{HMM}}(\theta,\Gamma;y) = \delta P(y_1;\theta) \prod_{t=2}^T \Gamma P(y_t;\theta) \mathbf{1}_N$$
%
where $\mathbf{1}_N$ is an $N$-dimensional column vector of ones and
%
$P(y_t;\theta)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry $f^{(i)}(y_t; \theta^{(i)})$.
%

@@

\subsection{The CarHMM and relaxing conditional independence}

A conditionally auto-regressive hidden Markov model, or CarHMM \citep{Lawler:2019}, is a generalization of an HMM which explicitly models auto-correlation in the observation sequence beyond the correlation induced by the hidden state process. Like a traditional HMM, a CarHMM is made up of a Markov chain of unobserved states $X_1,\ldots,X_T$ that can take on integer values between $1$ and $N$. It also has a transition probability matrix $\Gamma$ and initial distribution $\delta$ equal to the stationary distribution of $\Gamma$. Unlike a traditional HMM, the CarHMM assumes that the distribution of $Y_t$ conditioned on $X_1,\ldots, X_T$ and $Y_1,\ldots, Y_{t-1}$ depends on both $X_t$ and $Y_{t-1}$ rather than only $X_t$. The first emission $Y_1$ is treated as fixed as an initial value which does not depend upon $X_1$. We denote the conditional density or probability mass function of $Y_t$ given $Y_{t-1} = y_{t-1}$ and $X_t=i$ as $f^{(i)}(\cdot | y_{t-1}; \theta^{(i)})$ or simply $f^{(i)}(\cdot | y_{t-1})$. This model is highly general, as $f^{(i)}(\cdot | y_{t-1})$ can be any valid density or probability mass function that depends upon the parameters $\theta^{(i)}$ and the previous observation $Y_{t-1}$. 
As a concrete example, if $Y_t$ is a $K$-dimensional vector with components $Y_t(1),\ldots,Y_t(K)$, consider the component-wise sequences $\{Y_1(k), \ldots, Y_T(k)\}$ for $k = 1,\ldots K$. One may assume that these sequences are independent given $\{X_1,\ldots,X_T\}$ and that $Y_t(k)$ given $X_t = i$ is Normally distributed with parameters $\theta^{(i)}_k = \{\mu^{(i)}_k,\sigma^{(i)}_k,\phi^{(i)}_k\}$, where:
%
\begin{align}
\label{eqn:carhmm}
\begin{split}
\mathbb{E}(Y_{t}(k)|Y_{t-1}(k) = y_{t-1}(k),X_t=i) &= \phi^{(i)}_k ~ y_{t-1}(k) ~+ ~(1-\phi^{(i)}_k)  ~\mu^{(i)}_k, \\
\mathbb{V}(Y_t(k)| Y_{t-1}(k) = y_{t-1}(k), X_t = i) &= (\sigma^{(i)}_k)^2.
\end{split}
\end{align}
%
A CarHMM which follows Equation (\ref{eqn:carhmm}) can be viewed as a discrete time version of a state-switching Ornstein-Uhlenbeck process \citep{Michelot:2019}. This follows in the same way that an AR(1) process is the discrete-time version of a traditional Ornstein-Uhlenbeck process.

@@

\subsection{The HHMM and incorporating multiple scales}

A hierarchical hidden Markov model, or HHMM, accounts for processes occurring simultaneously at different scales by modelling both the coarse-scale process and fine-scale process with either HMMs \citep{Barajas:2017,Adam:2019} or CarHMMs. The coarse-scale model is as defined previously, where $X_1, \ldots, X_T$ make up an unobserved Markov chain with $N$ possible states and $Y_1,\ldots, Y_T$ are the corresponding observations with state-dependent parameters $\theta^{(i)}$ for $i = 1,\ldots,N$.   
%
In the hierarchical setting, however, each state $X_t$ also emits another sequence of fine-scale unobserved states, $X_t^* \equiv \{X_{t,1}^*,\ldots, X_{t,T_t^*}\}$, which in turn emits a sequence of fine-scale observations $Y_t^* \equiv \{Y_{t,1}^*,\ldots, Y_{t,T_t^*}\}$. For each $t$, the fine-scale process $\{X_t^*, Y_t^*\}$ then follows another HMM (or CarHMM) whose parameters depend on the value of $X_t$. If $X_t = i$, then the components of $X_t^*$ are a Markov chain with $N^{*(i)} \times N^{*(i)}$ transition probability matrix $\Gamma^{*(i)}$ and initial distribution $\delta^{*(i)}$ which we assume is equal to the stationary distribution of the chain. The distribution of $Y_{t,t^*}$ given $Y_{t,t^*-1} = y_{t,t^*-1}$, $X_{t,t^*}=i^*$, and $X_t=i$ is governed by the parameter $\theta^{*(i,i^*)}$ and has density or probability mass function denoted $f^{*(i,i^*)}\left(\cdot|y^*_{t,t^*-1}; \theta^{*(i,i^*)}\right)$ or simply $f^{*(i,i^*)}(\cdot|y^*_{t,t^*-1})$. We denote the set of fine-scale emission parameters corresponding to $X_t=i$ as $\theta^{*(i)}=\left\{\theta^{*(i,1)}, \ldots, \theta^{*(i,N^{*(i)})}\right\}$. In summary:
%
% this is new
%
\begin{gather*}
    \{Y, X\} \text{ follows a (Car)HMM with } \Gamma \in \bbR^{N \times N}, \\
    %
    \{Y_t | Y_{t-1} = y_{t-1}, X_t = i\} \sim f^{(i)}(\cdot|y_{t-1};\theta^{(i)}), \\
    %
    \{Y^*_t,X^*_t | X_t = i\} \text{ follows a (Car)HMM with } \Gamma^{*(i)} \in \bbR^{N^{*(i)} \times N^{*(i)}}, \\
    %
    \{Y^*_{t,t^*} | Y^*_{t,t^*-1} = y^*_{t,t^*-1}, X^*_{t,t^*} = i^*, X_t = i\} \sim f^{*(i,i^*)}(\cdot|y^{*}_{t,t^*-1};\theta^{(i,i^*)}).
\end{gather*}
Given the coarse-scale hidden states $X$, the $T+1$ sets $\{X_1^*, Y_1^*\}, \ldots, \{X_T^*, Y_T^*\}$ and $\{Y_1,\ldots,Y_T\}$ are assumed to be independent of one another.
%
% end new
%

@@

Due to the nested structure of the hierarchical hidden Markov model, the likelihood is easily calculated using the forward algorithm.
%
Let $y$ be the sequence of observed coarse-scale emissions and
$y^* \equiv \left\{y^*_1, \ldots,y^*_T\right\}$ be the collection of $T$ observed fine-scale emission vectors.
%
In addition, let $\theta^* \equiv \{\theta^{*(1)}, \ldots, \theta^{*(N)}\}$ denote the collection of all fine-scale emission parameters and $\Gamma^* \equiv \{\Gamma^{*(1)}, \ldots, \Gamma^{*(N)}\}$ denote the collection of all fine-scale transition probability matrices. The likelihood of the observed data is then
%
\begin{equation}
    \calL_{\text{HHMM}}(\theta,\theta^*,\Gamma,\Gamma^*;y,y^*) = \delta P(y_1,y_1^*;\theta,\theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P(y_t,y_t^*|y_{t-1};\theta,\theta^*,\Gamma^*) \mathbf{1}_N
    \label{eqn:HHMM_likelihood}
\end{equation}
%
where $P(y_1,y_1^*;\theta,\theta^*,\Gamma^*)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry equal to $f^{(i)}(y_1)\calL_{\text{(Car)HMM}}\left(\theta^{*(i)},
\Gamma^{*(i)};y_1^*\right)$ if the coarse-scale model is an HMM or $\calL_{\text{(Car)HMM}}\left(\theta^{*(i)},
\Gamma^{*(i)};y_1^*\right)$ if the coarse-scale model is a CarHMM. The fine-scale likelihood $\calL_{\text{(Car)HMM}}$ corresponds to the likelihood of the fine-scale model, which can be either a CarHMM or an HMM. For $t \geq 2$, $P(y_t,y_t^*|y_{t-1};\theta,\theta^*,\Gamma^*)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry equal to 
$f^{(i)}(y_t|y_{t-1})\calL_{\text{(Car)HMM}}\left(\theta^{*(i)},
\Gamma^{*(i)};y_t^*\right)$. Figure \ref{fig:models}c graphically displays the dependence structure for an HHMM.
%
%For more information on specific considerations for HHMMs, such as incorporating covariates into the probability transition matrix, state decoding, model selection and model checking, see \citet{Adam:2019}.

\subsection{The HMM-DFT and transforming fine-scale observations}
\label{subsec:STFT}

In many applications where data are collected at high frequencies, intricate dependency structures often arise within the fine-scale process which cannot be adequately modelled with any HMM-based model described thus far. 
%For example, the accelerometer data in Figure \ref{fig:data} sporadically displays clear periodic behaviour. 
To handle these additional fine-scale structures, we recommend replacing $Y_t^* = \{Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\}$ with relevant statistics that summarize any non-Markovian behaviour. To maintain the temporal structure of the fine-scale process, summary statistics can be calculated from a moving window over the elements $Y_t^*$. Subject matter experts are often required to determine the specific summary statistics employed as well as the optimal window size and stride length of the moving window. Larger stride lengths result in a loss of information but also reduce the dimension of the fine-scale process, which allows for faster model fitting. In addition, setting stride length and window size equal to one another avoids artificial residual correlation arising from overlapping windows. For our case study, we use the discrete Fourier transform (DFT) of a moving forward window of width $h$ and stride $h$ across $Y^*_t$. Namely:
%
\begin{align*}
    DFT\{Y^*_{t,t^*},\ldots, Y^*_{t,t^*+h-1}\}(k) = \sum_{n=0}^{h-1} Y^*_{t,t^*+n}\exp\left(-\frac{i 2\pi}{h} kn \right)
\end{align*}
%
for $t^* = 1,h+1,2h+1,\ldots$ and for $k = 0, 1, \ldots, h-1$. We omit the final window if $t^*+h-1$ exceeds $T^*_t$, and if $Y^*_{t,t^*}$ is a vector then the DFT is taken component-wise. We index each window with $\tilde{t}^*$ and calculate transformed observations $\Z_{t,\tilde{t}^*} \equiv \{\Zone_{t,\tilde{t}^*},\Ztwo_{t,\tilde{t}^*}\}$:
%
\begin{equation}
    \label{eqn:z}
    \Zone_{t,\tilde{t}^*} \equiv \frac{1}{h}\sum_{n=1}^{h}Y^*_{t,h(\tilde{t}^*-1)+n}, \quad \Ztwo_{t,\tilde{t}^*} \equiv \sum_{k=1}^{\tilde{\omega}}\Big|\Big|DFT\{Y^*_{t,h(\tilde{t}^*-1)+1},\ldots, Y^*_{t,h\tilde{t}^*}\}(k)\Big|\Big|^2,
\end{equation}
%
where $\tilde{\omega} \leq h-1$ is a problem-specific tuning parameter corresponding to the maximum recorded frequency within each window. In words, $\Zone_{t,\tilde{t}^*}$ is the average acceleration vector within window $\tilde{t}^*$ and $\Ztwo_{t,\tilde{t}^*}$ is the squared two-norm of the component of the window that can be attributed to frequencies between one and $\tilde{\omega}$ periods per window. %(by the Plancherel theorem). 
More intuitively, $\Ztwo_{t,\tilde{t}^*}$ corresponds to the ``wiggliness" of the acceleration data within dive $t$ and window $\tilde{t}^*$.

The fine-scale HMM must be redefined since $\Z_t$ exists on a coarser scale than $Y^*_t$ itself. In particular, if the length of $\Z_t$ is denoted as $\tilde{T}_t^*$ then there are only $\tilde{T}_t^*$ hidden states associated with $\Z_t = \big\{\Z_{t,1},\ldots, \Z_{t,\tilde{T}^*_t}\big\}$, which we denote as $\tilde{X}^*_t = \big\{\tilde{X}^*_{t,1},\ldots, \tilde{X}^*_{t,\tilde{T}^*_t}\big\}$. The fine-scale probability transition matrices  $\Gamma^{*(i)}$ and probability density functions $f^{*(i,i^*)}$ are then applied directly to $\tilde{X}^*_t$ and $\Z_t$ instead of $X^*_t$ and $Y^*_t$. The likelihood of this model is therefore identical to that of the original HMM, but $Y^*$ is replaced with $\Z$ and $X^*$ is replaced with $\tilde{X^*}$. In order to clearly differentiate models, we refer to an HMM with $\Z$ as observations and $\tilde{X^*}$ as hidden states as an HMM-DFT. Figure \ref{fig:models}d displays the dependence structure for an HMM-DFT.

@@

>>>>

Consider a sequence of $T$ curves, where any particular curve $t$ is characterized by a curve-level (or coarse-scale) observation $Y_t$ as well as a sequence of $T^*_t$ within-curve (or fine-scale) observations $Y^*_{t}$. Namely, $Y^*_{t} \equiv \big\{Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\big\}$ is made up of function values from curve $t$ and indexed using $t^*$. Both $Y_t$ and $Y_{t,t^*}$ can be either vectors or scalars. We call the sequence of coarse-scale observations $Y \equiv \big\{Y_1, \ldots, Y_T\big\}$ and the collection of all fine-scale observations $Y^* \equiv \big\{Y^*_1,\ldots,Y^*_T \big\}$. We first detail the structure of a traditional HMM followed by three variations which generalize its base structure. We then show how each of these generalized HMMs can be synthesized to form a wide variety of more complicated models.

@@

Using observation emissions, denoted here as $y \equiv \{y_1,\ldots,y_T\}$, we can find the maximum likelihood estimates of the parameters $\Gamma$ and $\theta \equiv \{\theta^{(1)},\ldots,\theta^{(N)}\}$. The likelihood $\calL_{\text{HMM}}$ can be evaluated using the well-known \textit{forward algorithm} \citep{Zucchini:2016}:
%
$$\calL_{\text{HMM}}(\theta,\Gamma;y) = \delta P(y_1;\theta) \prod_{t=2}^T \Gamma P(y_t;\theta) \mathbf{1}_N,$$
%
where $\mathbf{1}_N$ is an $N$-dimensional column vector of ones and
%
$P(y_t;\theta)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry $f^{(i)}(y_t; \theta^{(i)})$.
%

@@

\subsection{Relaxing conditional independence with the CarHMM}

A conditionally auto-regressive hidden Markov model, or CarHMM \citep{Lawler:2019}, is a generalization of an HMM which explicitly models auto-correlation in the observation sequence beyond the correlation induced by the hidden state process. Like a traditional HMM, a CarHMM is made up of a Markov chain of unobserved states $X_1,\ldots,X_T$ that can take on integer values between $1$ and $N$. A CarHMM also has a transition probability matrix $\Gamma$ and initial distribution $\delta$ equal to the stationary distribution of $\Gamma$. Unlike a traditional HMM, the CarHMM assumes that the distribution of $Y_t$ conditioned on $X_1,\ldots, X_T$ and $Y_1,\ldots, Y_{t-1}$ depends on both $X_t$ and $Y_{t-1}$ rather than only $X_t$. The first emission $Y_1$ is treated as a fixed initial value which does not depend upon $X_1$. We denote the conditional density or probability mass function of $Y_t$ given $Y_{t-1} = y_{t-1}$ and $X_t=i$ as $f^{(i)}(\cdot | y_{t-1}; \theta^{(i)})$ or simply $f^{(i)}(\cdot | y_{t-1})$. This model is highly general, as $f^{(i)}(\cdot | y_{t-1})$ can be any valid density or probability mass function that depends upon the parameters $\theta^{(i)}$ and the previous observation $Y_{t-1}$. 
As a concrete example, if $Y_t$ is a scalar, then one may assume that $Y_t$ given $X_t = i$ is Normally distributed with parameters $\theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$, where:
%
\begin{align}
\label{eqn:carhmm}
\begin{split}
\mathbb{E}(Y_{t}|Y_{t-1} = y_{t-1},X_t=i) &= \phi^{(i)} ~ y_{t-1} ~+ ~(1-\phi^{(i)})  ~\mu^{(i)}, \\
\mathbb{V}(Y_t| Y_{t-1} = y_{t-1},X_t=i) &= (\sigma^{(i)})^2.
\end{split}
\end{align}
%
A CarHMM which follows Equation (\ref{eqn:carhmm}) can be viewed as a discrete time version of a state-switching Ornstein-Uhlenbeck process \citep{Michelot:2019}. This follows in the same way that an AR(1) process is the discrete-time version of a traditional Ornstein-Uhlenbeck process.

@@

\subsection{Incorporating multiple scales with the HHMM}

A hierarchical hidden Markov model, or HHMM, accounts for processes occurring simultaneously at different scales by modelling both the coarse-scale process and fine-scale process with either HMMs \citep{Barajas:2017,Adam:2019} or CarHMMs. The coarse-scale model is as defined previously, where $X_1, \ldots, X_T$ make up an unobserved Markov chain with $N$ possible states and $Y_1,\ldots, Y_T$ are the corresponding observations with state-dependent parameters $\theta^{(i)}$ for $i = 1,\ldots,N$.   
%
In the hierarchical setting, however, each state $X_t$ also emits another sequence of fine-scale unobserved states, $X_t^* \equiv \{X_{t,1}^*,\ldots, X_{t,T_t^*}\}$, which in turn emits a sequence of fine-scale observations $Y_t^* \equiv \{Y_{t,1}^*,\ldots, Y_{t,T_t^*}\}$. For each $t$, the fine-scale process $\{X_t^*, Y_t^*\}$ then follows another HMM (or CarHMM) whose parameters depend on the value of $X_t$. If $X_t = i$, then the components of $X_t^*$ are a Markov chain with $N^{*(i)} \times N^{*(i)}$ transition probability matrix $\Gamma^{*(i)}$ and initial distribution $\delta^{*(i)}$ which we assume is equal to the stationary distribution of the chain. The distribution of $Y_{t,t^*}$ given $Y_{t,t^*-1} = y_{t,t^*-1}$, $X_{t,t^*}=i^*$, and $X_t=i$ is governed by the parameter $\theta^{*(i,i^*)}$ and has density or probability mass function denoted $f^{*(i,i^*)}\left(\cdot|y^*_{t,t^*-1}; \theta^{*(i,i^*)}\right)$ or simply $f^{*(i,i^*)}(\cdot|y^*_{t,t^*-1})$. We denote the set of fine-scale emission parameters corresponding to $X_t=i$ as $\theta^{*(i)}=\left\{\theta^{*(i,1)}, \ldots, \theta^{*\left(i,N^{*(i)}\right)}\right\}$. In summary:
%
% this is new
%
\begin{gather*}
    \{Y, X\} \text{ follows a (Car)HMM with } \Gamma \in \bbR^{N \times N}, \\
    %
    \{Y_t | Y_{t-1} = y_{t-1}, X_t = i\} \sim f^{(i)}(\cdot|y_{t-1};\theta^{(i)}), \\
    %
    \{Y^*_t,X^*_t | X_t = i\} \text{ follows a (Car)HMM with } \Gamma^{*(i)} \in \bbR^{N^{*(i)} \times N^{*(i)}}, \\
    %
    \{Y^*_{t,t^*} | Y^*_{t,t^*-1} = y^*_{t,t^*-1}, X^*_{t,t^*} = i^*, X_t = i\} \sim f^{*(i,i^*)}(\cdot|y^{*}_{t,t^*-1};\theta^{(i,i^*)}).
\end{gather*}
Given the coarse-scale hidden states $X$, the $T+1$ sets $\{X_1^*, Y_1^*\}, \ldots, \{X_T^*, Y_T^*\}$ and $\{Y_1,\ldots,Y_T\}$ are assumed to be independent of one another.
%
% end new
%

@@

Due to the nested structure of the HHMM, the likelihood is easily calculated using the forward algorithm.
%
Let $y$ be the sequence of observed coarse-scale emissions and
$y^* \equiv \left\{y^*_1, \ldots,y^*_T\right\}$ be the collection of $T$ observed fine-scale emission vectors.
%
In addition, let $\theta^* \equiv \{\theta^{*(1)}, \ldots, \theta^{*(N)}\}$ denote the collection of all fine-scale emission parameters and $\Gamma^* \equiv \{\Gamma^{*(1)}, \ldots, \Gamma^{*(N)}\}$ denote the collection of all fine-scale transition probability matrices. The likelihood of the observed data is then
%
\begin{equation}
    \calL_{\text{HHMM}}(\theta,\theta^*,\Gamma,\Gamma^*;y,y^*) = \delta P(y_1,y_1^*;\theta,\theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P(y_t,y_t^*|y_{t-1};\theta,\theta^*,\Gamma^*) \mathbf{1}_N
    \label{eqn:HHMM_likelihood}
\end{equation}
%
where $P(y_1,y_1^*;\theta,\theta^*,\Gamma^*)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry equal to $f^{(i)}(y_1)\calL_{\text{(Car)HMM}}\left(\theta^{*(i)},
\Gamma^{*(i)};y_1^*\right)$ if the coarse-scale model is an HMM or $\calL_{\text{(Car)HMM}}\left(\theta^{*(i)},
\Gamma^{*(i)};y_1^*\right)$ if the coarse-scale model is a CarHMM. The fine-scale likelihood $\calL_{\text{(Car)HMM}}$ corresponds to the likelihood of the fine-scale model, which can be either a CarHMM or an HMM. For $t \geq 2$, $P(y_t,y_t^*|y_{t-1};\theta,\theta^*,\Gamma^*)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry equal to 
$f^{(i)}(y_t|y_{t-1})\calL_{\text{(Car)HMM}}\left(\theta^{*(i)},
\Gamma^{*(i)};y_t^*\right)$. Figure \ref{fig:models}c graphically displays the dependence structure for an HHMM.
%
%For more information on specific considerations for HHMMs, such as incorporating covariates into the probability transition matrix, state decoding, model selection and model checking, see \citet{Adam:2019}.

\subsection{Transforming fine-scale observations with the HMM-DFT}
\label{subsec:STFT}

In many applications where data are collected at high frequencies, intricate dependency structures often arise within the fine-scale process which cannot be adequately modelled with any HMM-based model described thus far. 
%For example, the accelerometer data in Figure \ref{fig:data} sporadically displays clear periodic behaviour. 
To handle these additional fine-scale structures, we recommend replacing $Y_t^* = \{Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\}$ with relevant statistics that summarize any non-Markovian behaviour. To maintain the temporal structure of the fine-scale process, summary statistics can be calculated from a moving window over the elements $Y_t^*$. Subject matter experts are often required to determine the specific summary statistics employed as well as the optimal window size and stride length of the moving window. Larger stride lengths result in a loss of information but also reduce the dimension of the fine-scale process, which allows for faster model fitting. In addition, setting stride length and window size equal to one another avoids artificial residual correlation arising from overlapping windows. For our case study, we use the discrete Fourier transform (DFT) of a moving forward window of width $h$ and stride $h$ across $Y^*_t$. Namely:
%
\begin{align*}
    DFT\{Y^*_{t,t^*},\ldots, Y^*_{t,t^*+h-1}\}(k) = \sum_{n=0}^{h-1} Y^*_{t,t^*+n}\exp\left(-\frac{i 2\pi}{h} kn \right)
\end{align*}
%
for $t^* = 1,h+1,2h+1,\ldots$ and $k = 0, 1, \ldots, h-1$ with $i^2 = -1$. We omit the final window if $t^*+h-1$ exceeds $T^*_t$, and if $Y^*_{t,t^*}$ is a vector then the DFT is taken component-wise. We index each window with $\tilde{t}^*$ and calculate transformed observations $\Z_{t,\tilde{t}^*} \equiv \{\Zone_{t,\tilde{t}^*},\Ztwo_{t,\tilde{t}^*}\}$:
%
\begin{equation}
    \label{eqn:z}
    \Zone_{t,\tilde{t}^*} \equiv \frac{1}{h}\sum_{n=1}^{h}Y^*_{t,h(\tilde{t}^*-1)+n}, \quad \Ztwo_{t,\tilde{t}^*} \equiv \sum_{k=1}^{\tilde{\omega}}\Big|\Big|DFT\{Y^*_{t,h(\tilde{t}^*-1)+1},\ldots, Y^*_{t,h\tilde{t}^*}\}(k)\Big|\Big|^2,
\end{equation}
%
where $\tilde{\omega} \leq h-1$ is a problem-specific tuning parameter corresponding to the maximum recorded frequency within each window. In words, $\Zone_{t,\tilde{t}^*}$ is the average value of $Y^*_t$ within window $\tilde{t}^*$ and $\Ztwo_{t,\tilde{t}^*}$ is the squared two-norm of the component of the window that can be attributed to frequencies between one and $\tilde{\omega}$ periods per window. %(by the Plancherel theorem). 
More intuitively, $\Ztwo_{t,\tilde{t}^*}$ corresponds to the ``wiggliness" of the fine-scale data within curve $t$ and window $\tilde{t}^*$.

The fine-scale HMM must be redefined since $\Z_t$ exists on a coarser scale than $Y^*_t$ itself. In particular, if the length of $\Z_t$ is denoted as $\tilde{T}_t^*$ then there are only $\tilde{T}_t^*$ hidden states associated with $\Z_t = \big\{\Z_{t,1},\ldots, \Z_{t,\tilde{T}^*_t}\big\}$, which we denote as $\tilde{X}^*_t = \big\{\tilde{X}^*_{t,1},\ldots, \tilde{X}^*_{t,\tilde{T}^*_t}\big\}$. The fine-scale probability transition matrices  $\Gamma^{*(i)}$ and probability density functions $f^{*(i,i^*)}$ are then applied directly to $\tilde{X}^*_t$ and $\Z_t$ instead of $X^*_t$ and $Y^*_t$. The likelihood of this model is therefore identical to that of the original HMM, but $Y^*$ is replaced with $\Z$ and $X^*$ is replaced with $\tilde{X^*}$. To clearly differentiate models, we refer to an HMM with $\Z$ as observations and $\tilde{X^*}$ as hidden states as an HMM-DFT. Figure \ref{fig:models}d displays the dependence structure for an HMM-DFT.

@@

%the fine scale behaviour follows a continuous time HMM.  The parameters of the fine scale distributions depend on hidden states.  However the states are modelled as components in a mixture distribution, not on an HMM.
